Comparing Facebook's Prophet Forecasting to an ARIMA Model: A Diebold-Mariano Evaluation of the Top40 Index

### 2017-10-14 template/Draft_Paper.Rmd
Draft paper edits: Data plot fixed, table 1 and 2 changed to markdown tables, Some data plots were printed directly from markdown (rest of the code will be added to markdown doc later)

###Quasi-Bayesian methodss incorporating Analysts' knowlegde

The stock price forecasting approaches considered so far are purely apply technical data analysis methods to generate stock price forecasting models. However, studies that have compared pure technical approaches to those that incorporate the analysts knowledge in a quasi-Bayesian form have been shown to be superior at predicting market prices [@givoly1984quality \& @guerard1989combining]. The first instance of combining time-series model forecasts and an analysts forecasts to obtain superior forecasts of a stocks annual earnings can be dated to 1989. @guerard1989combining used an additive model to combine consensus security analyst forecasts from the S&P Annual Earnings forecaster and annual earnings forecasts generated by an ARIMA model with a constant. The combined model that was estimated using ordinary least squares reduced the mean square error of the time series and analysts forecasts from 1.28 and 1.27, respectively to 1.04. These results suggest that analysts can substantially reduce forecasting errors by combining both approaches [@guerard1989combining]. @zahedi2015application applied the same overarching principle when they used principle component analysis to determine an appropriate input variable, which they then used to train an ANN in an attempt to predict stock prices on the Tehran Stock Exchange. Their model yielded acceptable forecasting errors and performed better than the pure ANN which further reiterates the potential of incorporating an analysts knowledge to improve stock price forecasts, in both linear and nonlinear models.

###Diebold-Mariano Evaluation
As is evident from the previous sections, the modelling approach and assumptions of Prophet forecasting-at-scale differ to those of an ARIMA model. Since the two methods differ, the Diebold-Mariano evaluation (DM Test) is preferred as a means of testing forecasting accuracy. This is because the DM Test is a model free test of forecasting accuracy. It is applicable to a wide range of situations from multi-period forecasts, non-Gaussian forecast residuals, non-quadratic loss functions and even serially correlated data[@Mariano2000].

@@

The ARIMA model is a generalisation of Autoregressive Moving Average (ARMA) models which combine autoregression and moving average features [@box1970time]. In the context of returns, the ARIMA formulation for the return at time index $t$, $r_t$, is given by,
$$  r_t = \phi_1 r_{t-1} + \phi_2 r_{t-2} + ...+ \phi_p r_{t-p} + \epsilon_t +
        \psi_1 \epsilon_{t-1} + \psi_2 \epsilon_{t-2} + ... +\psi_q \epsilon_{t-q} $$  
where:  
$\bullet \phi_i$ and $\phi_j$ are the parameters to be estimated;  
$\bullet r_{t-k}$ is the $k^{th}$ lagged return;   
$\bullet \epsilon_t$ is the error term at time $t$ which is assumed to be white noise and  
$\bullet\epsilon_{t-k}$ is the $k^{th}$ lagged error.   
$\bullet$p is the order of the autoregressive component (dependenceon history of $r_t$)
$\bullet$q is the order of the moving average component (dependenceon history of $\epsilon_t$)

>>>>

###Quasi-Bayesian methods incorporating Analysts' knowledge

The stock price forecasting approaches considered so far purely apply technical data analysis methods to generate stock price forecasting models. However, studies that have compared pure technical approaches to those that incorporate the analysts knowledge in a quasi-Bayesian form have been shown to be superior at predicting market prices [@givoly1984quality \& @guerard1989combining]. The first instance of combining time-series model forecasts and an analysts forecasts to obtain superior forecasts of a stocks annual earnings can be dated to 1989. @guerard1989combining used an additive model to combine consensus security analyst forecasts from the S&P Annual Earnings forecaster and annual earnings forecasts generated by an ARIMA model with a constant. The combined model that was estimated using ordinary least squares reduced the mean square error of the time series and analysts forecasts from 1.28 and 1.27, respectively to 1.04. These results suggest that analysts can substantially reduce forecasting errors by combining both approaches [@guerard1989combining]. @zahedi2015application applied the same overarching principle when they used principle component analysis to determine an appropriate input variable, which they then used to train an ANN in an attempt to predict stock prices on the Tehran Stock Exchange. Their model yielded acceptable forecasting errors and performed better than the pure ANN which further reiterates the potential of incorporating an analysts knowledge to improve stock price forecasts, in both linear and nonlinear models.

###Diebold-Mariano Evaluation
As is evident from the previous sections, the modelling approach and assumptions of Prophet differ to those of an ARIMA model. Since the two methods differ, the Diebold-Mariano evaluation (DM Test) is preferred as a means of testing forecasting accuracy. This is because the DM Test is a model free test of forecasting accuracy. It is applicable to a wide range of situations from multi-period forecasts, non-Gaussian forecast residuals, non-quadratic loss functions and even serially correlated data[@Mariano2000].

@@

The ARIMA model is a generalisation of Autoregressive Moving Average (ARMA) models which combine autoregression and moving average features [@box1970time]. In the context of returns, the ARIMA formulation for the return at time index $t$, $r_t$, is given by,
$$  r_t = \phi_1 r_{t-1} + \phi_2 r_{t-2} + ...+ \phi_p r_{t-p} + \epsilon_t +
        \psi_1 \epsilon_{t-1} + \psi_2 \epsilon_{t-2} + ... +\psi_q \epsilon_{t-q} $$  
where:  
$\bullet \phi_i$ and $\psi_j$ are the parameters to be estimated;  
$\bullet r_{t-k}$ is the $k^{th}$ lagged return;   
$\bullet \epsilon_t$ is the error term at time $t$ which is assumed to be white noise and  
$\bullet\epsilon_{t-k}$ is the $k^{th}$ lagged error.   
$\bullet$p is the order of the autoregressive component (dependenceon history of $r_t$)
$\bullet$q is the order of the moving average component (dependenceon history of $\epsilon_t$)

### 2017-10-14 template/Draft_Paper.tex
Draft paper edits: Data plot fixed, table 1 and 2 changed to markdown tables, Some data plots were printed directly from markdown (rest of the code will be added to markdown doc later)

Artificial Neural Networks (ANNs) are a multi-layered perceptron as
illustrated in figure 1.\\
\includegraphics{ANN.png}\\
\emph{Figure 1 : Illustration of an Artificial neural network} (Tanikic
and Despotovic 2012)\\
An ANN consists of a sorted triple \((K, A, \omega)\). \(K\) is a set
representing the multiple layers/levels of the network i.e.~input,
hidden and output levels. A is a set of pairs that represents the value
of the index of the estimate in the \(i^{th}\) node of the \(k^{th}\)
layer of the network. In the enlarged section of Figure 1,
\(a_{i,k} \in A\). Lastly, \(\omega\) represents a function that defines
the weights (\(\omega_{i,r}\)) of connections between neurons (Kriesel
2007). Artificial Neural networks apply iterated optimization of model
parameters across the network in the form of weights conditional on
observed values to learn (Segaran 2007). Several studies have compared
ARIMA model forecasts to ANN and their conclusions are contradictory.

@@

\subsubsection{Quasi-Bayesian methodss incorporating Analysts'
knowlegde}\label{quasi-bayesian-methodss-incorporating-analysts-knowlegde}

The stock price forecasting approaches considered so far are purely
apply technical data analysis methods to generate stock price
forecasting models. However, studies that have compared pure technical
approaches to those that incorporate the analysts knowledge in a
quasi-Bayesian form have been shown to be superior at predicting market
prices (Givoly and Lakonishok 1984 \& Guerard (1989)). The first
instance of combining time-series model forecasts and an analysts
forecasts to obtain superior forecasts of a stocks annual earnings can
be dated to 1989. Guerard (1989) used an additive model to combine
consensus security analyst forecasts from the S\&P Annual Earnings
forecaster and annual earnings forecasts generated by an ARIMA model
with a constant. The combined model that was estimated using ordinary
least squares reduced the mean square error of the time series and
analysts forecasts from 1.28 and 1.27, respectively to 1.04. These
results suggest that analysts can substantially reduce forecasting
errors by combining both approaches (Guerard 1989). Zahedi and Rounaghi
(2015) applied the same overarching principle when they used principle
component analysis to determine an appropriate input variable, which
they then used to train an ANN in an attempt to predict stock prices on
the Tehran Stock Exchange. Their model yielded acceptable forecasting
errors and performed better than the pure ANN which further reiterates
the potential of incorporating an analysts knowledge to improve stock
price forecasts, in both linear and nonlinear models.

@@

As is evident from the previous sections, the modelling approach and
assumptions of Prophet forecasting-at-scale differ to those of an ARIMA
model. Since the two methods differ, the Diebold-Mariano evaluation (DM
Test) is preferred as a means of testing forecasting accuracy. This is
because the DM Test is a model free test of forecasting accuracy. It is
applicable to a wide range of situations from multi-period forecasts,
non-Gaussian forecast residuals, non-quadratic loss functions and even
serially correlated data(Mariano 2000).

@@

The ARIMA model is a generalisation of Autoregressive Moving Average
(ARMA) models which combine autoregression and moving average features
(Box and Jenkins 1970). In the context of returns, the ARIMA formulation
for the return at time index \(t\), \(r_t\), is given by,
\[  r_t = \phi_1 r_{t-1} + \phi_2 r_{t-2} + ...+ \phi_p r_{t-p} + \epsilon_t +
        \psi_1 \epsilon_{t-1} + \psi_2 \epsilon_{t-2} + ... +\psi_q \epsilon_{t-q} \]\\
where:\\
\(\bullet \phi_i\) and \(\phi_j\) are the parameters to be estimated;\\
\(\bullet r_{t-k}\) is the \(k^{th}\) lagged return;\\
\(\bullet \epsilon_t\) is the error term at time \(t\) which is assumed
to be white noise and\\
\(\bullet\epsilon_{t-k}\) is the \(k^{th}\) lagged error.\\
\(\bullet\)p is the order of the autoregressive component (dependenceon
history of \(r_t\)) \(\bullet\)q is the order of the moving average
component (dependenceon history of \(\epsilon_t\))

@@

The data set used throughout this paper comprises of the daily closing
prices of the equally-weighted FTSE/JSE Top40 price index for the five
year period from 31/07/2012 to 30/07/2017. This equates to a total of
1766 trading days excluding weekends and holidays. This period was
chosen for its recency and to avoid exposing the results to the noise
associated with the high volatility experienced in the stock market
during the 2008 financial crisis. The aim is to compare the predictive
accuracy of the two models under standard market conditions before other
confounding factors can be closely analysed. Figure 3 shows the actual
realisations of the price index data which was sourced from Datastream.

\includegraphics[width=1.00000\textwidth]{plot of Price index.png}\\
\emph{Figure 3: Realisation of the FTSE/JSE Top40 Price Index over the
period of study}

@@

In this paper, the \textbf{log return}(which will be used
interchangeably with \textbf{return}), over the period \([t,t+1]\) is
denoted by \(r_t\), and was calculated by evaluating;\\
\[
r_t = log(\frac{P_{t+1}}{P_t})  \label{eq1}
\]\\
Where \(P_t\) and \(P_{t+1}\) represent the price of the index at time
\(t\) and \(t+1\) respectively.

@@

Observations from the above plots informed the choice of candidate
\emph{ARIMA(p,d,q)} models to fit to the data. The candidate models that
were fit to the in-sample data are shown in Table 1 with their
respective AIC values. The candidate model with the lowest AIC is the
\(ARIMA(2,0,2)\). This model was thus selected as the ARIMA forecasting
model.\\
\includegraphics[width=1.00000\textwidth]{AIC.png}\\
\emph{Table 1: AIC for Candidate ARIMA models}

@@

>>>>

Artificial Neural Networks (ANNs) are a multi-layered perceptron as
illustrated in figure 1.\\
\includegraphics{ANN.png}\\
\emph{Figure 1 : Illustration of an Artificial neural network} (Tanikic
and Despotovic 2012)\\
An ANN consists of a sorted triple \((K, A, \omega)\). \(K\) is a set
representing the multiple layers/levels of the network i.e.~input,
hidden and output levels. A is a set of pairs that represents index for
the estimate in the \(i^{th}\) node of the \(k^{th}\) layer of the
network. In the enlarged section of Figure 1, \(a_{i,k} \in A\). Lastly,
\(\omega\) represents a function that defines the weights
(\(\omega_{i,r}\)) of connections between neurons/nodes (Kriesel 2007).
Artificial Neural networks apply iterated optimization of model
parameters across the network in the form of weights conditional on
observed values to learn (Segaran 2007). Several studies have compared
ARIMA model forecasts to ANN and their conclusions are contradictory.

@@

\subsubsection{Quasi-Bayesian methods incorporating Analysts'
knowledge}\label{quasi-bayesian-methods-incorporating-analysts-knowledge}

The stock price forecasting approaches considered so far purely apply
technical data analysis methods to generate stock price forecasting
models. However, studies that have compared pure technical approaches to
those that incorporate the analysts knowledge in a quasi-Bayesian form
have been shown to be superior at predicting market prices (Givoly and
Lakonishok 1984 \& Guerard (1989)). The first instance of combining
time-series model forecasts and an analysts forecasts to obtain superior
forecasts of a stocks annual earnings can be dated to 1989. Guerard
(1989) used an additive model to combine consensus security analyst
forecasts from the S\&P Annual Earnings forecaster and annual earnings
forecasts generated by an ARIMA model with a constant. The combined
model that was estimated using ordinary least squares reduced the mean
square error of the time series and analysts forecasts from 1.28 and
1.27, respectively to 1.04. These results suggest that analysts can
substantially reduce forecasting errors by combining both approaches
(Guerard 1989). Zahedi and Rounaghi (2015) applied the same overarching
principle when they used principle component analysis to determine an
appropriate input variable, which they then used to train an ANN in an
attempt to predict stock prices on the Tehran Stock Exchange. Their
model yielded acceptable forecasting errors and performed better than
the pure ANN which further reiterates the potential of incorporating an
analysts knowledge to improve stock price forecasts, in both linear and
nonlinear models.

@@

As is evident from the previous sections, the modelling approach and
assumptions of Prophet differ to those of an ARIMA model. Since the two
methods differ, the Diebold-Mariano evaluation (DM Test) is preferred as
a means of testing forecasting accuracy. This is because the DM Test is
a model free test of forecasting accuracy. It is applicable to a wide
range of situations from multi-period forecasts, non-Gaussian forecast
residuals, non-quadratic loss functions and even serially correlated
data(Mariano 2000).

@@

The ARIMA model is a generalisation of Autoregressive Moving Average
(ARMA) models which combine autoregression and moving average features
(Box and Jenkins 1970). In the context of returns, the ARIMA formulation
for the return at time index \(t\), \(r_t\), is given by,
\[  r_t = \phi_1 r_{t-1} + \phi_2 r_{t-2} + ...+ \phi_p r_{t-p} + \epsilon_t +
        \psi_1 \epsilon_{t-1} + \psi_2 \epsilon_{t-2} + ... +\psi_q \epsilon_{t-q} \]\\
where:\\
\(\bullet \phi_i\) and \(\psi_j\) are the parameters to be estimated;\\
\(\bullet r_{t-k}\) is the \(k^{th}\) lagged return;\\
\(\bullet \epsilon_t\) is the error term at time \(t\) which is assumed
to be white noise and\\
\(\bullet\epsilon_{t-k}\) is the \(k^{th}\) lagged error.\\
\(\bullet\)p is the order of the autoregressive component (dependenceon
history of \(r_t\)) \(\bullet\)q is the order of the moving average
component (dependenceon history of \(\epsilon_t\))

@@

The data set used throughout this paper comprises of the daily closing
prices of the FTSE/JSE Top40 price index for the five year period from
31/07/2012 to 31/07/2017. This equates to a total of 1305 trading days
excluding weekends and holidays. This period was chosen for its recency
and to avoid exposing the results to the noise associated with the high
volatility experienced in the stock market during the 2008 financial
crisis. The aim is to compare the predictive accuracy of the two models
under standard market conditions before other confounding factors can be
closely analysed. Figure 3 shows the actual realisations of the price
index data which was sourced from Datastream.

\includegraphics{Draft_Paper_files/figure-latex/unnamed-chunk-1-1.pdf}
\emph{Figure 3: Realisation of the FTSE/JSE Top40 Price Index over the
period of study}

@@

In this paper, the \textbf{log return} (which will be used
interchangeably with \textbf{return}), over the period \([t,t+1]\) is
denoted by \(r_t\), and was calculated by evaluating;\\
\[
r_t = log(\frac{P_{t+1}}{P_t})  \label{eq1}
\]\\
Where \(P_t\) and \(P_{t+1}\) represent the price of the index at time
\(t\) and \(t+1\) respectively.

@@

Observations from the above plots informed the choice of candidate
\emph{ARIMA(p,d,q)} models to fit to the data. The candidate models that
were fit to the in-sample data are shown in Table 1 with their
respective AIC values. The candidate model with the lowest AIC is the
\(ARIMA(2,0,2)\). This model was thus selected as the ARIMA forecasting
model.

\begin{table}[H]
\centering
\begin{tabular}{ll}
  \hline
ARIMA(p,d,q) model & AIC \\ 
  \hline
(0,0,0) & -8804 \\ 
  (2,0,0) & -8808 \\ 
  (0,0,2) & -8808 \\ 
  (2,0,2) & -8817 \\ 
   \hline
\end{tabular}
\caption{AIC for Candidate ARIMA models} 
\end{table}

@@

\begin{table}[H]
\centering
\begin{tabular}{lllll}
  \hline
Forecast Horizon, h & Forecast Model &  ME  & RMSE & MAE \\ 
  \hline
1 & ARIMA & -0.0004110 & 0.0103 & 0.007718 \\ 
    & Prophet & 0.0000176 & 0.0104 & 0.007749 \\ 
  5 & ARIMA & -0.0003635 & 0.0103 & 0.007692 \\ 
    & Prophet & 0.004445 & 0.0103 & 0.007709 \\ 
  20 & ARIMA & -0.0001769 & 0.0102 & 0.007649 \\ 
    & Prophet & -0.0000406 & 0.0102 & 0.007660 \\ 
   \hline
\end{tabular}
\caption{Forecast error diagnostic statistics over different forecasting horizons} 
\end{table}
